Neural Networks for Neural Tumors\\ \vspace{-0.6em} {\footnotesize Final Project for \textit{ECS 171 - Machine Learning} at UC Davis, with Professor Setareh Rafatirad}

### 2025-06-05 latex/draft.tex
more drafting, finished revising methodology and experimental results

\large 
\section{\large Introduction (Complete)}
As a team of five young computer science students with no prior machine learning experience, we ambitously decided to tackle the real world challenge of brain tumor classification. Through hours and hours of research and trial and error, we learned a lot, and we have a lot to share.

@@

\subsection{\large Approach}
A classification machine learning model is a great approach to this problem. Using a convolutional neural network and an image dataset, we can train a model to accuractely classify brain tumors. With its complex structure, a CNN can detect nuances and patterns in the scans that are otherwise hard to spot.

@@

% Both figures in one row
% TODO: Move these to be under EDA in the right spots
\begin{figure*}[!ht]
    \centering
    \includegraphics[width=0.49\textwidth]{ImagesPerClassTraining.png}
    \hfill
    \includegraphics[width=0.49\textwidth]{ImagesPerClassTesting.png}
    \caption{\large The dataset has a consistent number of training images across classes, except for "no tumor". This should not be an issue because "no tumor" imagery has much less variance. For testing, the distribution of images is more varied, but this does not affect model performance, since the model is not trained on this data.}
    \label{fig:images-per-class}
\end{figure*}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=4in]{ImageDimensions.png}
    \caption{\large It is clear to see the images in the dataset are of many varying dimensions. We do see higher densities around the 200x200 and 500x500 dimensions, though. Regardless, such variation means image preprocessing and resizing is necessary before training.}
    \label{Dimensions of images in the dataset}
\end{figure*}

Taking a look at Figure 2, we can see that the dataset is comprised of images of many varying sizes. There is a concentration around the 200x200 and 500x500 sizes. With this information, we know we will have to perform some image pre-processing to convert all images into the same size so the model has consistent input. We will provide our CNN with the conventional 224x244 image dimensions.

\section{\large Methodology (1-2 pages)}

\subsection{\large Feature Selection}
As mentioned in our dataset description, we don't have labels for our input features like a numerical dataset would have. We only have our images, which are technically each a set of pixel intensity values, but it means we can't select certain features to train our dataset on. Instead, we can perform some image preprocessing and hyperparameter tuning to increase our model's accuracy.

Looking at Figure 3 again, we see all of our images have different dimensions. As a preprocessing step, we will resize them all to be 224x224 pixels, so that our model has consistent input.  The resizing step is not as simple as it sounds though, we must be careful because shrinking or enlarging our images can create unwanted noise in our data. After imagine resizing, we then normalize each pixel's value from its 3-channel RGB form.

Our training dataset consists of 2870 MRI scans. If we want to give our model more images to train on, we can perform data augmentation like mentioned previously. We can rotate, flip, and zoom our images to create 'artificial' data that will hopefully prevent overfitting and allow for our model to capture underlying patterns better. We also hope this will allow for our model to predict against a larger variety of testing data (i.e. rotated images).

The hyperparameters of our model are crucial to achieving a good fit. In our first iteration of the model, some of the hyperparameters we chose were as follows:
%TODO --- fix table
- Image size: 256x256 pixels		- Batch size: 32		- Epochs: 30	
- Loss:  Categorical Cross Entropy 	- Metrics: Accuracy		- Dropout Rate: 0.25 and 0.50
We plan on fine tuning these parameters with techniques like grid search as we continue to optimize our model.

@@

Our first early prototype was a CNN model with 4 convolutional blocks, followed by fully connected layers using the Keras Sequential API. We trained the model over 30 epochs, with each epoch giving us higher accuracy on the training data, peaking at 93\% after the last epoch. However, when performing predictions on our testing data, the model saw a big dip in performance. With an average accuracy of 42\%, our model was heavily overfitting on the training data.

% @Nam is this true? k-fold CV etc.?
To improve upon our first model, we applied k-fold cross validation and tuned hyperparameters such as dropout rate, number of layers/neurons, etc. We implemented grid search to tune our hyperparameters. We also planned to use transfer learning to reduce the number of learned parameters and implicitly filter irrelevant features, which can greatly reduce overfitting risks. Going forward, we'd use PyTorch, which gave us more flexibility in controlling forward pass, backpropagation, and optimization, allowing us to develop a better CNN model.

\section{\large Experimental Results (2 pages)}

Since deep learning is very efficient in detecting features from an image, and this characteristic is also beneficial to building other models using different base algorithms, we have mainly focused on developing a Convolutional Neural Network (CNN) model with a high accuracy by iteratively refining both model architecture and training strategy. This section outlines the chronological development and iterative experimentation we have applied to finalize the set of models. While the detailed performance of each model will be elaborated on in the following sections, this discussion emphasizes the architecture refinement, technical adjustments, and discoveries made during the project that improved the overall performance over time.

@@

Our very first model was built from scratch using TensorFlow’s Sequential API. The architecture consisted of 4 convolution blocks with subsequent increases in the number of filters (kernels), 
each of which is followed by a max pooling layer helping reduce the spatial dimensions. The last layers of the model are characterized by a dense layer of 128 neurons with ReLU activation and an output layer of 4 neurons with Softmax activation for multi-class classification, each of which is accompanied by a dropout layer to mitigate overfitting.

This model was trained over 20 epochs with categorical cross-entropy as the loss function and the Adam optimizer. 
The model ended up with a training accuracy of 92.56\% and a validation accuracy of 76.67\%. When tested on the testing set, the model only gave an accuracy of 42.13\%. The significant gap between training and testing performance implies signs of overfitting.

@@

In the next progression, we decided to employ the transfer learning technique, using the ResNet50 model, which is pre-trained on the ImageNet dataset. Instead of training the entire neural network from scratch, we leveraged the architecture of ResNet50 as a feature extractor by eliminating its output layer and freezing all other layers. The classification head was built on top of the pre-trained model, which served as a convolutional base, followed by a Global Average Pooling layer to compress spatial dimensions. The classifier was completed by a fully connected dense layer of 128 neurons with ReLU activation and an output layer with softmax activation, which was also accompanied by a dropout rate of 0.3 to prevent overfitting. \

The model was trained over 20 epochs. This transition yielded a significant improvement in validation accuracy (89.20\% compared to 76.67\% in the first model) while maintaining a reasonable training accuracy (92.56\%). Motivated by this encouragement, we tried to experiment with selective fine-tuning the model by unfreezing the last 30 layers of the ResNet50 pre-trained model, allowing the weights of those layers to be adjusted during the backpropagation. With this technique, this model ended up with an accuracy of 97.53\%, 89.20\% and 79.19\% for the training, validation, and testing sets, respectively. At this stage, we also introduced new training practices such as learning rate management and early stopping, which shaped our training strategy for later experiments

@@

Motivated by the significant improvement in the previous stage, we had defined that transfer learning would be one of the most critical techniques in developing our CNN model. We decided to try different pre-trained models, from simpler models such as VGG16, VGG19 to more complicated ones in ResNet and EfficientNet families. Even with either layer-wise fine-tuning or standard fine-tuning techniques applied to those models, we had noticed the common trait, which is a persistent underperformance on the testing test (with an accuracy around 70.00\% to 80\%) despite high accuracy in validation metrics (with an accuracy around 90\% to 93\%). This discrepancy prompted us to re-evaluate our dataset. Initially, we had used the dataset split provided by Kaggle. We suspected that there might have been an imbalanced presentation across classes or feature distribution in the default split. To resolve this issue, we decided to recombine the training and testing sets with shuffle, then split the entire dataset with a ratio consistent with that of the default split. \

With available computational resources, we also upgraded to ResNet101, a larger architecture compared to ResNet50, and transitioned to Keras Functional API to implement our model's architecture for better flexibility and reusability. We employed the same structure as in the ResNet50 setup. The model was initially trained over 20 epochs with all ResNet101 layers frozen, yielding a training accuracy of 91.72\% and validation accuracy of 85.07\%. When the fine-tuning technique was applied (with the last 40 layers unfrozen), the model reached 99.67\% training accuracy and 95.83\% validation accuracy after being trained over 4 additional epochs. The model also performed much better on the testing data, yielding 96.94\% accuracy with reasonably high f1-scores for each class label.

@@

To develop other classifiers that incorporate algorithms rather than deep learning, we decided to employ Support Vector Machine (SVM) and Random Forest Classifier (RF) models. We leveraged the fine-tuned CNN model as the feature extraction for our selected dataset in order to obtain high accuracy while keeping the training process relatively simple. More specifically, we extracted the output of the GlobalAveragePooling2D layer of the fine-tuned CNN model, which includes 2048-dimensional feature vectors, and that result would serve as the input for each non-deep learning model. In other words, our SVM and RF models were trained on a more semantically meaningful input space compared to raw pixels.

After conducting hyperparameter tuning, we obtained an overall accuracy of 95.66\% and 94.90\% for SVM and RF classifiers, respectively. Although these two models were less efficient in distinguishing between glioma and meningioma compared to the CNN model, other performance metrics were relatively high. In general, these results demonstrate that classical classifiers, when paired with feature extractors, can deliver highly competitive performance with significantly reduced training complexity

>>>>

\large 
\section{\large Introduction (Complete)}

As a team of five young computer science students with no prior machine learning experience, we ambitously decided to tackle the real world challenge of brain tumor classification. Through hours and hours of research and trial and error, we learned a lot, and we have a lot to share.

@@

\subsection{\large Approach}

A classification machine learning model is a great approach to this problem. Using a convolutional neural network and an image dataset, we can train a model to accuractely classify brain tumors. With its complex structure, a CNN can detect nuances and patterns in the scans that are otherwise hard to spot.

@@

% Both figures in one row
% TODO: Move these to be under EDA in the right spots
\begin{figure*}[!ht]
    \centering
    \includegraphics[width=0.49\textwidth]{ImagesPerClassTraining.png}
    \hfill
    \includegraphics[width=0.49\textwidth]{ImagesPerClassTesting.png}
    \caption{\large The dataset has a consistent number of training images across classes, except for "no tumor". This should not be an issue because "no tumor" imagery has much less variance. For testing, the distribution of images is more varied, but this does not affect model performance, since the model is not trained on this data.}
    \label{images-per-class}
\end{figure*}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=5in]{ImageDimensions.png}
    \caption{\large It is clear to see the images in the dataset are of many varying dimensions. We do see higher densities around the 200x200 and 500x500 dimensions, though. Regardless, such variation means image preprocessing and resizing is necessary before training.}
    \label{Dimensions of images in the dataset}
\end{figure*}

Taking a look at Figure 2, we can see that the dataset is comprised of images of many varying sizes. There is a concentration around the 200x200 and 500x500 sizes. With this information, we know we will have to perform some image pre-processing to convert all images into the same size so the model has consistent input.

\section{\large Methodology (Complete)}

\subsection{\large Feature Selection}

As mentioned in the dataset description, we don't have labels for our input features like a numerical dataset would have. We only have our images, which are technically each a set of pixel intensity values, but it means we can't select certain features to train our model on. Instead, we can perform some image preprocessing and hyperparameter tuning to increase our model's accuracy.

Looking at Figure 2 again, we see all of our images have different dimensions. As a preprocessing step, we will resize all images to be 224x224 pixels, the conventional CNN input size. The resizing step is not as simple as it sounds though, we must be careful not to create unwanted noise in our data. After imagine resizing, we then normalize each pixel's value from its 3-channel RGB form.

The training dataset consists of 2870 MRI scans. To increase the size of the training dataset, we can perform data augmentation like mentioned previously. We can rotate, flip, and zoom the images to create `artificial' data that would hopefully reduce any overfitting and allow for the model to better capture underlying patterns. We also would expect this technique to allow for the model to predict against a larger variety of testing data (i.e. rotated images).

Along with input features, the hyperparameters of a model are crucial to achieving a good fit. The key hyperparameters for the CNN model were: batch size for gradient descent, training epochs, loss function, dropout rate, and a few others. We used grid search to find optimal parameters for all three of our models, more on that later.

@@

Our first prototype model was a CNN with four convolutional blocks, followed by fully connected layers using the Keras Sequential API. We trained the model over 20 epochs, with each epoch giving us higher accuracy on the training data, peaking at 93\%. However, when performing predictions on our testing data, the model saw a big dip in performance. With an average accuracy of 42\%, we knew immediately that our model was heavily overfitting the training data.

To improve upon the first model, we applied k-fold cross validation and tuned hyperparameters such as dropout rate, number of layers, number of neurons, etc. This was the first time we used grid search to tune hyperparameters. It was also at this stage we decided that going forward, we would use transfer learning to reduce the number of learned parameters and implicitly filter irrelevant `features'. Transfer learning, which is the process of leveraging a pre-trained model to improve your own, can greatly reduce overfitting risks. From this point on, we used PyTorch. Pytorch gave us more flexibility in controlling forward pass, backpropagation, and optimization, allowing us to develop a better-fitting CNN model.

\section{\large Experimental Results (Complete)}

Neural networks like CNNs are very good at detecting features and learning patterns in data. Because of this, we can leverage deep learning to preprocess data, and then use that data to train other models. So far, this report has focused on developing a CNN model with high accuracy by iteratively refining both model architecture and training strategy. While CNNs are the most common algorithm choice for image classification, they are not the only option. Image classification can be done by various other algorithms, including support vector machine (SVM), and random forest (RF).

While the detailed performance of each model we created will be discussed in the following sections, this section outlines the chronological development and iterative experimentation that lead us to our final models.

@@

Like mentioned before, our very first model was built from scratch using TensorFlow's Sequential API. The model consisted of four convolution blocks with subsequent increases in the number of filters (kernels). Each block was followed by a max pooling layer helping reduce the spatial dimensionality. Additionally, each hidden layer of the model was accompanied by a dropout layer to mitigate overfitting. The last hidden layer of the model was a dense layer of 128 neurons using the ReLU activation function. The output layer was four neurons using Softmax as the activation function. Softmax, given its consistent output between 0 and 1, is the standard choice of activation function for the output layer in multi-class classification.

This model was trained over 20 epochs using the Adam optimizer and categorical cross-entropy as the loss function. And as mentioned before, the model ended up with a training accuracy of 93\% and a testing accuracy of 42\%, implying significant overfitting.

@@

In our next step, we decided to employ transfer learning. We started with the ResNet50 model, a pre-trained model trained on the ImageNet dataset. We leveraged the architecture of ResNet50 as a feature extractor by eliminating its output layer and freezing all other layers. The classification head was built on top of the pre-trained model, which served as a convolutional base. The base was followed by a Global Average Pooling layer to compress spatial dimensions. The classifier was again completed by a fully connected dense layer of 128 neurons with the ReLU activation function and a Softmax output layer. The layer of 128 neurons was accompanied by a dropout rate of 0.3 to help prevent overfitting.

The model was trained over 20 epochs. The transition to transfer learning yielded a significant improvement in validation accuracy (89\% compared to 76\% in the first model) while maintaining a reasonable training accuracy of 93\%. Motivated by this encouragement, we experimented with selective fine-tuning the model by unfreezing the last 30 layers of the ResNet50 pre-trained model, allowing the weights of those layers to be adjusted during backpropagation. With this adjustment, the model ended up with a training accuracy of 97\%, and a testing accuracy of 79\%. Although the model was still slightly overfitting, we had made huge improvements. At this stage, we also introduced new training practices such as the learning rate hyperparameter and early stopping, which shaped our training strategy for later experiments.

@@

Motivated by the significant improvement in the previous stage, we had defined that transfer learning would be one of the most critical techniques in developing our CNN model. We decided to try different pre-trained models, from simpler models such as VGG16 and VGG19, to more complex ones in the ResNet and EfficientNet families. Even with either layer-wise fine-tuning or standard fine-tuning techniques applied to those models, we had noticed a common trait, persistent underperformance on the testing test (with an accuracy around 70\% to 80\%). This discrepancy prompted us to re-evaluate our dataset. Initially, we had used the dataset split provided by Kaggle. We suspected that there might have been an imbalanced presentation across classes or feature distribution in the default split. To resolve this issue, we decided to recombine the training and testing sets with shuffle, then split the entire dataset with a ratio consistent with that of the default split.

With limited available computational resources, we also upgraded our pre-trained model to ResNet101, a more complex model compared to ResNet50. We employed the same structure as in the ResNet50 setup. The model was initially trained over 20 epochs with all ResNet101 layers frozen, yielding a training accuracy of 92\% and validation accuracy of 85\%. When the fine-tuning technique was applied (with the last 40 layers unfrozen), the model reached 99\% training accuracy and 96\% validation accuracy after being trained over four additional epochs. The model also performed much better on the testing data, yielding 97\% accuracy with reasonably high f1-scores for each class label. This was our breakthrough.

@@

To develop other classifiers that incorporate algorithms rather than deep learning, we decided to employ Support Vector Machine (SVM) and Random Forest (RF) models. We leveraged the fine-tuned CNN model as the feature extraction for our selected dataset in order to obtain high accuracy while keeping the training process relatively simple. More specifically, we extracted the output of the GlobalAveragePooling2D layer of the fine-tuned CNN model, which includes 2048-dimensional feature vectors, and that result would serve as the input for each non-deep learning model. In other words, our SVM and RF models were trained on a more semantically meaningful input space compared to raw pixels.

After conducting hyperparameter tuning, we obtained an overall accuracy of 96\% and 95\% for SVM and RF classifiers respectively. In general, these results demonstrate that classical classifiers, when paired with feature extractors, can deliver highly competitive performance for image classification, with significantly reduced training complexity.
